A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-04 latex/main.tex
Updates from Overleaf

\documentclass[12pt,letterpaper]{article}
\usepackage[a4paper, total={7in, 10in}]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}
\usepackage{helvet}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{orcidlink}
\usepackage[super,comma,sort&compress]
   {natbib}\bibliographystyle{numbered}
\usepackage[right]{lineno} \linenumbers
\usepackage[normalem]{ulem}

@@

The LS HMM at the heart of sc2ts provides a simple and powerful means
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution
for the putative recombinant sequence
with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination.
% TODO
[FIXME need to insert a sentence saying this is similar to
metrics used by RIPPLES and other methods, contrasting
recombinant vs non-recombinant explanations.]
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

@@

Figure~\ref{fig:recombinant_evidence}A plots
the number of mutations required
in the recombinant versus non-recombinant
Viterbi solutions (upper scatter plot)
and the resulting distribution of the number of
averted mutations (lower histogram)
over all 354 recombination events that pass QC.
Also shown is the 3SEQ classification status of these recombinants.
The 5\% of recombinants that do not pass 3SEQ validation are
among the most weakly supported, with 4 or 5 averted mutations.
[TODO discuss B and C]
Figures~\ref{fig:recombinant_evidence_cov_recomb} and
~\ref{fig:recombinant_evidence_rebar} 
plot the same information for CovRecomb and rebar validation,
showing that their validation status tends to be function
of the divergence between the recombinant parents.

% SHZ: We should be more careful with the wording.
% Recombination is likely frequent,
% but detectable recombination
% leading to lineages that are competitive enough to be sampled
% compared to their (VOC) parents
% is rare.
Recombination is rare in SARS-CoV-2, and
ultimately judgments about plausibility require expert analysis.
To aid such interpretation and help identify QC issues,
we developed a visualization tool
inspired by SnpIt~\cite{OToole2024} and 
the recombinant view of RIVET~\cite{Smith2023tracking},
showing the recombination-informative sites
between a recombinant and its parent sequences, and
highlighting potentially problematic loci.
Each copying pattern shows the positions
where the allelic state
in a recombinant (middle row) matches that of
the left (upper row, colored green) 
or right parent (lower row, colored blue),
or where the recombination event requires
a \textit{de novo} mutation (gold, with mutational change below).
[FIXME: Should we also mention the red/orange lines
under the patterns that identify adjacent sites?]
Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
illustrating a spectrum of plausibility.
XA is an example of a highly plausible recombinant,
shown by the large number of supporting loci on the left and right.
As we have only one \textit{de novo} mutation,
we can be sure that the parents are closely observed
[FIXME: Some details about XA here, how well the parents were sampled, etc.].
XBB is also strongly supported by many loci on the left and right, 
but the presence of many \textit{de novo} mutations
observed in the copying patterns suggests the existence of
many genetically distinct, un-sampled ancestors
[FIXME: in this case, probably BJ.1, see issue 336],
and some caution should be exercised when interpreting them.
See Figures~\ref{fig:recombinant_evidence}D and
\ref{fig:pango_x_copying_patterns}
for copying patterns of the Pango X lineages, and
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problems
(and more details on the visualizations).

@@

To compare the performance of sc2ts and UShER+RIPPLES 
in terms of recombinant detection, we ran RIPPLES on the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
Note that we are interested in the direct output of sc2ts and UShER+RIPPLES
here, and do not perform QC filtering of either.
Using the RIPPLES parameter $p=4$ (equivalent to $k=4$ used 
to build the sc2ts ARG) resulted in 1174 reported recombination events.
We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
were classified as recombinant, giving an overall validation rate of 51\%.
Aligning the events reported by both methods using the sets of descendant
samples (STAR Methods), 468 of the events are shared by both,
with the intersection comprising 40\% of the RIPPLES events
and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination of two factors.
Firstly, 408 of these RIPPLES identified recombination events
have one or more parents that are \emph{younger} than the
recombinant sequence based on metadata. Thus, the parental
sequences were simply not available to match against for 
sc2ts (which enforces strict time ordering) when the 
recombinant sequence was inserted. Secondly, of the 179
remaining events with matching clades and in which the 
sc2ts node has more than 4 mutations, of the total
1,167 mutations, 452 of these occurred at sites that were 
remapped during post-hoc parsimony. The remaining ``missing''
events are likely attributable to the finer details of the 
differences in tree building approaches between UShER and sc2ts.

@@

>>>>

\documentclass[12pt,letterpaper]{article}
\usepackage[a4paper, total={7in, 10in}]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}
\usepackage{helvet}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{placeins}
\usepackage{orcidlink}
\usepackage[super,comma,sort&compress]
   {natbib}\bibliographystyle{numbered}
\usepackage[right]{lineno} \linenumbers
\usepackage[normalem]{ulem}

@@

The LS HMM at the heart of sc2ts provides a simple and powerful means
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution
for the putative recombinant sequence
with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination (similar to the  
``parsimony improvement'' score in RIPPLES~\cite{Turakhia2022}).
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

@@

Figure~\ref{fig:recombinant_evidence}A plots
the number of mutations required
in the recombinant versus non-recombinant
Viterbi solutions (upper scatter plot)
and the resulting distribution of the number of
averted mutations (lower histogram)
over all 354 recombination events that pass QC.
Also shown is the 3SEQ classification status of these recombinants.
The 5\% of recombinants that do not pass 3SEQ validation are
among the most weakly supported, with 4 or 5 averted mutations.
Figures~\ref{fig:recombinant_evidence}B and C shows the 
distribution of recombination events as a function the 
divergence between the parents.
Figures~\ref{fig:recombinant_evidence_cov_recomb} and
~\ref{fig:recombinant_evidence_rebar} 
plot the same information as Figure~\ref{fig:recombinant_evidence} 
broken down by CovRecomb and rebar validation status.

Detectable recombination is rare in SARS-CoV-2, and
ultimately judgments about plausibility require expert analysis.
To aid such interpretation and help identify QC issues,
we developed a visualization tool
inspired by SnpIt~\cite{OToole2024} and 
the recombinant view of RIVET~\cite{Smith2023tracking},
showing the recombination-informative sites
between a recombinant and its parent sequences, and
highlighting potentially problematic loci.
Each copying pattern shows the positions
where the allelic state
in a recombinant (middle row) matches that of
the left (upper row, colored green) 
or right parent (lower row, colored blue),
or where the recombination event requires
a de novo mutation (gold, with mutational change below).
Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
illustrating a spectrum of plausibility.
XA is an example of a highly plausible recombinant,
shown by the large number of supporting loci on the left and right.
XBB is also strongly supported by many loci on the left and right, 
but the presence of many de novo mutations
observed in the copying patterns suggests the existence of
many genetically distinct, un-sampled ancestors
and some caution should be exercised when interpreting them
(see Document~\ref{sec:dd_xbb} for more details).
See Figures~\ref{fig:recombinant_evidence}D and
\ref{fig:pango_x_copying_patterns}
for copying patterns of the Pango X lineages, and
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problems
(and more details on the visualizations).

@@

To compare the performance of sc2ts and UShER+RIPPLES 
in terms of recombinant detection, we ran RIPPLES on the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
Note that we are interested in the direct output of sc2ts and UShER+RIPPLES
here, and do not perform QC filtering of either.
Using the RIPPLES parameter $p=4$ (matching $k=4$ in the sc2ts ARG) resulted in 1174 reported recombination events.
We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
were classified as recombinant, giving an overall validation rate of 51\%.
Aligning the events reported by both methods using the sets of descendant
samples (STAR Methods), 468 of the events are shared by both,
with the intersection comprising 40\% of the RIPPLES events
and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination of two factors.
Firstly, 408 of these RIPPLES identified recombination events
have one or more parents that are \emph{younger} than the
recombinant sequence based on metadata. Thus, the parental
sequences were simply not available to match against for 
sc2ts (which enforces strict time ordering) when the 
recombinant sequence was inserted. Secondly, of the 179
remaining events with matching clades and in which the 
sc2ts node has more than 4 mutations, of the total
1,167 mutations, 452 of these occurred at sites that were 
remapped during post-hoc parsimony. The remaining ``missing''
events are likely attributable to the finer details of the 
differences in tree building approaches between UShER and sc2ts.

@@

\FloatBarrier
